---
output:
  github_document:
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# march.madness

<!-- badges: start -->
<!-- badges: end -->

The goal of march.madness is to help you fill out your march madness bracket. This package will help you make informed decisions on who to pick.

## Installation

You can install the development version of march.madness from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("djperrenoud/march.madness")
```

## Example

To select a single team to view their season stats and use in functions:

```{r team}
library(march.madness)
selected_team <- team_select("Utah St.", 2024)
selected_team
```

You can also select a matchup of two teams to compare:

```{r matchup}
selected_matchup <- matchup_select("Utah St.", 2024, "Purdue", 2024)
selected_matchup
```

To visualize the season summary of a team:

```{r tables, eval=FALSE}
team_table(selected_team)

stat_table(selected_team)
```

To visualize the matchup of two teams:

```{r tables2, eval=FALSE}
matchup_chart(selected_matchup)
```

To view the power of each team and the probability of each team winning:

```{r power}
matchup_prob(selected_matchup)
```
