---
title: "March Madness Project Vignette"
author: "AJ Haygood, Dylan Perrenoud, Jonah Feigleson"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
vignette: "%\\VignetteIndexEntry{March Madness Project Vignette} %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}\n"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Github URL

https://github.com/djperrenoud/march.madness

# Introduction

This vignette describes how to use the `march.madness` package. This package 
provides tools for predicting outcomes of NCAA March Madness basketball games 
based on historical and current season data.The functions in this package can 
be used to query the KenPom database for relevant basketball statistics for 
college men's basketball teams from the years 2008-2024. Additionally, they 
can be used to output  These tools are useful for a number of reasons. 
informative tables and graphs for individual teams as well as matchups. The 
matchup probability function predicts the probability of winning for two teams 
if they played each other. 

These functions are useful for a couple of situations. First, they allow users 
to make data-informed decisions when predicting outcomes of March Madness 
tournament games. This year, 17.3 million people created March Madness brackets 
for men's college basketball on ESPN's Tournament Challenge platform. Having 
tools available to predict win probabilities and compare important stats, like 
those in this package, would help people with little background knowledge about 
basketball stay competitive in bracket challenges. Second, it gives basketball 
fans a tool to analyze their favorite team's winning ability. The functions in 
this package can be used to compare the same basketball team from different 
years. This allows users to assess the progression of their teams year to year.

The multi-year comparison capabilities of our package set it apart from other
tools for March Madness bracket building. Additionally, our matchup probability 
function is robust. After testing it on multiple 2024 March Madness brackets, it
predicted at least one of the teams in the national championship correctly, and 
predicted 9 out of 13 upsets. 

This vignette explains the importance of each function in our package, and how 
they work together to help users prepare March Madness brackets and make 
user-friendly comparisons of college basketball teams. 


# Installation

You can install the released version of `march.madness` with:

```{r, eval=FALSE}
remotes::install_github("djperrenoud/march.madness")
```
```{r, eval=TRUE}
library(march.madness)
```

# Data Cleaning

This section outlines the data cleaning procedures implemented in the 
'clean_data.R' script, which are designed to refine the data set in alignment 
with the project's objectives. The process involves removing columns that are 
not essential for the analysis. Specifically, non-essential data such as 
conference identifiers, team IDs, and ranking values of various metrics are 
excluded. For example, while the Kenpom adjusted offensive value is retained, 
its corresponding ranking values are discarded. This approach enables the direct
utilization of actual metric values for more effective comparisons and 
predictions, rather than relying on their ranked positions.

# Main Functions

## Selection Functions

### Team Name Formatting

#### `format_team_name(team)`

This formatting function improves user experience by accounting for alternate 
spellings for team names. For example, if a user were to use the input 
"Utah St." in the team_select function, we understand which team they would like
information on. However, if the KenPom dataset has the team listed as Utah 
State, the team_select function would throw an error. This function standardizes
the user's inputs with the names used in the dataset. It runs the user's team 
name through various cases listed below, and then coerces the inputted name to 
its correct format. Additionally, if a user forgot to include common 
typographical elements like capitalization and punctuation, the function 
accounts for these as well. This function is necessary for a smooth user 
experience, and is called in the team_select and matchup_select functions to 
minimize easily avoided errors. To understand the specifics of how the function 
operates, refer to the details section below.

-   **Details**:

    -   Takes user input an lower cases every letter from the start

    -   First checks if the input is part of the "special cases" to insure 
    correct capitalization:

        -   Loyola MD
        -   LIU Brooklyn
        -   UT Arlington
        -   UNC Asheville
        -   McNeese St.
        -   UC Davis
        -   UC Irvine
        -   UC Santa Barbara

    -   Checks if the input is apart of the "saint cases" to output "Saint" 
    correctly:

        -   Mount St. Mary's is the exception
        -   Saint Louis
        -   Saint Joseph's
        -   Saint Mary's
        -   Saint Peter's

    -   Checks if the school is an A&M or A&T school:

        -   North Carolina A&T
        -   Praire View A&M
        -   Texas A&M
        -   Texas A&M Corpus Christi

    -   Checks if the input is apart of the 4 letter all caps cases:

        -   UNLV
        -   UTEP
        -   UTSA
        -   UMBC

    -   Checks if the input is the "Miami case"

        -   Since there are two universities named "Miami", if a user inputs 
        "Miami" it automatically will switch the input to "Miami FL" due to the 
        fact that Miami Ohio is not included in the dataset. The user will 
        receive a warning letting them know that "Miami" was coerced to 
        "Miami FL".

    -   Checks if the input is apart of the 3 letter all caps cases

    -   After checking these cases, general capitalization rules will apply, 
    capitalizing the first letter in each word

    -   The function also allows for the user to input "St." as "state" or "st" 
    to allow some error from the user.

-   **Example**:

``` {r, include = TRUE}
# Example of handling a common abbreviation
formatted_team <- format_team_name("utah st")
print(formatted_team) # Outputs "Utah St."

# Handling a university name with special abbreviations
formatted_team <- format_team_name("uc irvine")
print(formatted_team) # Outputs "UC Irvine"

# Correcting and standardizing 'Saint' school names
formatted_team <- format_team_name("st joseph's")
print(formatted_team) # Outputs "Saint Joseph's"
```

### Team Selection

#### `team_select(team, year)`

The team_select and matchup_select functions are the backbone of the package. 
Team_select is the user's way to access the KenPom dataset, and convert it to an
output that can be used by the other functions in the package. To use the 
team_select function, the user passes a team name and a year to the function. 
The function searches through a "cleaned" version of the dataset and returns a 
data frame containing 45-50 relevant statistics describing that iteration of the 
team. These dataframes are then converted into descriptive tables by 
subsequent functions in the package. The subsetted data frames are returned as 
custom S3 objects of class 'teamdata'. While these data frames may be of 
interest to package users who are familiar with all the statistics we found to 
be relevant to our other functions, the 'teamdata' objects are primarily useful 
to other functions in the package, which we will discuss later on. 

-   **Example**:

``` {r, include = TRUE}
selected_team <- team_select("Baylor", 2024)
print(selected_team)
```

### Match-up Selection

#### `matchup_select(team1, year1, team2, year2)`

The matchup_select function is very similar to the team_select function, with a 
few key differences. This function takes two teams as inputs, and returns a 
data frame that contains the team stats for both teams. While the team_select 
function is used to make tables for individual teams, the matchup_select 
function is used to make comparisons between two teams. These data frames are 
returned as S3 objects of class "teamdata_matchup". These frames are useful
to users because they format the data to be called by other functions. Other 
functions use these "teamdata_matchup" objects to calculate which team is most 
likely to win in a matchup, as well as side-by-side comparisons of stats that 
are crucial to team success. These functions will be explained in the following 
sections.

-   **Example**:

``` {r, include = TRUE}
selected_matchup <- matchup_select("Baylor", 2023, "Purdue", 2023)
print(selected_matchup)
```

## Visualization Functions

### Team Table

#### `team_table(selected_team)`

This is the first of our table functions, and helps users understand basic but 
useful information about March Madness teams. These tables are particularly 
useful for users who are new to college basketball. For instance, if a user was 
making a bracket, or even watching a college basketball game, and was unfamiliar 
with UNC, they could use the team_table function to learn about it. They could 
call `team_table(team_select("UNC", 2024))`, and be presented with a table
explaining the team's conference, seed, wins and losses. The table also includes
the round they made it to that year (assuming that year's tournament has already 
been played). This information helps the user to understand very quickly how 
successful a team has been, and will help them make decisions on how far they 
think the team will go in current or future tournaments.

-   **Example**:

``` {r, include = TRUE}
team_table(team_select("Baylor", 2024))
```

### Stat Table

#### `stat_table(selected_team)`

This is the second of the package's table functions, and is useful for 
understanding more in-depth information on a given basketball team. The tables 
returned by the function include percentages for four key statistics: 
efficiency field goals, free throw rate, offensive rebounds, and turnover rates. 
These stats act as key indicators for how well a team is playing. If a team is 
has high percentages in these categories, even if they are seeded lower, they 
may have a good chance of upsetting higher ranked teams. This function is useful 
for users who are faced with close matchups in their brackets, and want help to 
decide if they should call an upset or not. It also can be used to familiarize 
themselves with why teams are seeded where they are, outside of just wins and 
losses.The example below illustrates how the function is called in tandem with 
the team_select function. 

-   **Example**:

``` {r, include = TRUE}
stat_table(team_select("Baylor", 2024))
```

### Match up Visualization

#### `matchup_chart(teamdata_matchup)`

The matchup_chart function helps users visualize team matchups. The chart takes 
a "teamdata_matchup" object and returns a side-by-side bar chart comparing the 
four key indicator statistics returned by the stat_table function. After using 
the team_table and stat_table functions for multiple teams, a user can use the 
matchup_chart function to see a comparison of teams they have familiarized 
themselves with. This information helps the users make a decision on who to pick
as a winner for close matchups. They can also use this function to compare the 
same team over two different years to understand if the program is getting 
better or worse.

-   **Example**:

``` {r, include = TRUE}
matchup_chart(matchup_select("Baylor", 2024, "Purdue", 2024))
```

## Match up Function

#### `matchup_prob(teamdata_matchup)`

This probability function gives the most accessible and useful matchup analysis 
for users. It takes a "teamdata_matchup" object, and uses its contained values 
to calculate a "power" rating and return a winning probability for each team. 
Once a user has familiarized themselves with two teams using the table and chart 
functions, they can call the matchup_prob function to get number value that 
shows how likely each team is to win a matchup. A user could use this function 
for each matchup in their bracket to pick a winner, making bracket building less
stressful. This function can also give users peace of mind knowing their picks 
are backed by observed sports statistics. While this function isn't a perfect 
predictor, it will give users a good shot at winning bracket challenges. 

-   **Example**:

``` {r, include = TRUE}
teamdata_matchup <- matchup_select("Purdue", 2024, "Utah St.", 2024)
matchup_prob(teamdata_matchup)
```

# Conclusion

The march.madness package described above provides user-friendly analytical 
tools for basketball fans of all levels to design brackets and make March 
Madness predictions. For a user who is a social basketball viewer, the table and 
chart functions will help them familiarize themselves with relevant college 
basketball teams. The information returned from these functions will not only 
help people build informed brackets, but can help people participate in everyday 
conversations about college basketball. The probability function can help people 
make fast judgements on who is going to win a matchup, especially when the 
seeds are close. Additionally, the stat_table and matchup_chart functions can 
help explain why some teams are on unexpected winning streaks based on key 
statistics from the KenPom dataset. All of these functions can also be used to 
compare the same team year-to-year, which sets this package apart from other 
predictive tools. This functionality can show how a team is progressing, and 
even help predict how good a team will be in future years. And while this 
package may not win you millions of dollars, it might just win you clout in your 
family bracket challenge. 
